Calendar-based graphics for visualizing people's daily schedules

### 2018-10-08 main.tex
should fix up the chinese fonts issue in adobe
fixed up the bib reference
reformatted to R Journal style

\documentclass[article]{jss}
\usepackage[utf8]{inputenc}

\providecommand{\tightlist}{%
  \setlength{\itemsep}{0pt}\setlength{\parskip}{0pt}}

\author{
Earo Wang\\Monash University \And Dianne Cook\\Monash University \And Rob J Hyndman\\Monash University
}
\title{Calendar-based graphics for visualizing people's daily schedules}

\Plainauthor{Earo Wang, Dianne Cook, Rob J Hyndman}
\Plaintitle{Calendar-based graphics for visualizing people's daily schedules}

\Abstract{
This paper describes a \code{frame\_calendar} function that organizes
and displays temporal data, collected on sub-daily resolution, into a
calendar layout. Calendars are broadly used in society to display
temporal information, and events. The \code{frame\_calendar} uses linear
algebra on the date variable to create the layout. It utilizes the
grammar of graphics to create the plots inside each cell, and thus
synchronizes neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior in Melbourne, Australia,
based on counts which are captured at hourly intervals by sensors
scattered around the city. Faceting by the usual features such as day
and month, was insufficient to examine the behavior. Making displays on
a monthly calendar format helps to understand pedestrian patterns
relative to events such as work days, weekends, holidays, and special
events. The layout algorithm has several format options and variations.
It is implemented in the \proglang{R} package \pkg{sugrrants}.
}

\Keywords{data visualization, statistical graphics, time series, grammar of graphics, \proglang{R}}
\Plainkeywords{data visualization, statistical graphics, time series, grammar of graphics, R}

%% publication information
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{}
%% \Acceptdate{2012-06-04}

\Address{
    Earo Wang\\
  Monash University\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{earo.wang@monash.edu}\\

      Dianne Cook\\
  Monash University\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{dicook@monash.edu}\\

      Rob J Hyndman\\
  Monash University\\
  Department of Econometrics and Business Statistics, Monash University,
  VIC 3800 Australia\\
  E-mail: \email{rob.hyndman@monash.edu}\\

% Pandoc header

\usepackage{amsmath}

\usepackage{amsthm}
\newtheorem{theorem}{Theorem}[section]
\newtheorem{lemma}{Lemma}[section]
\theoremstyle{definition}
\newtheorem{definition}{Definition}[section]
\newtheorem{corollary}{Corollary}[section]
\newtheorem{proposition}{Proposition}[section]
\theoremstyle{definition}
\newtheorem{example}{Example}[section]
\theoremstyle{definition}
\newtheorem{exercise}{Exercise}[section]
\theoremstyle{remark}
\newtheorem*{remark}{Remark}
\newtheorem*{solution}{Solution}
\begin{document}

\section{Introduction}\label{introduction}

We develop a method for organizing and visualizing temporal data,
collected at sub-daily intervals, into a calendar layout. The calendar
format is created using linear algebra, fitting nicely into the
\pkg{tidyverse} \citep{R-tidyverse} suite. In particular, the resulting
coordinates synchronize with \pkg{ggplot2} \citep{R-ggplot2} graphics
for exploring data. The purpose of the calendar-based visualization is
to provide insights into people's daily schedules relative to events
such as work days, weekends, holidays, and special events. This work was
originally motivated by studying foot traffic in the city of Melbourne,
Australia \citep{ped}. There have been 43 sensors installed that count
pedestrians every hour across the downtown area (Figure
\ref{fig:ped-map}). The dataset can shed light on people's daily
rhythms, and assist the city administration and local businesses with
event planning and operational management. A routine examination of the
data would involve constructing conventional time series plots to catch
a glimpse of temporal patterns. The faceted plots in Figure
\ref{fig:time-series-plot}, give an overall picture of the foot traffic
at 3 different sensors over 2016. Further faceting by day of the week
(Figure \ref{fig:facet-time}) provides a better glimpse of the daily and
sub-daily pedestrian patterns. The sensor data, like many temporal
datasets on human behaviors, lends itself to a number of exploratory
data visualization challenges:

@@

\begin{CodeChunk}
\begin{figure}

@@

\caption[Map of the Melbourne city area with purple dots
indicating sensor locations.]{Map of the Melbourne city area with purple dots
indicating sensor locations.}\label{fig:ped-map}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

@@

\caption[Time series plots showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Colored by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.]{Time series plots showing the number of
pedestrians in 2016 measured at 3 different sensors in the city of
Melbourne. Colored by the sensors, small multiples of lines show that
the foot traffic varies from one sensor to another in terms of both time
and number. The weekly patterns look distinctive across these 3 sensors.
There is an eye-catching spike which occurred at the State Library,
caused by the annual White Night event on 20th of February.}\label{fig:time-series-plot}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

@@

\caption[Hourly pedestrian counts for 2016 faceted by
sensors and days of the week using lines. It features at least two types
of seasons---time of day and day of week---across all the sensors. The
White Night effect on increasing pedestrian counts can be seen on
Saturday in February at the State Library.]{Hourly pedestrian counts for 2016 faceted by
sensors and days of the week using lines. It features at least two types
of seasons---time of day and day of week---across all the sensors. The
White Night effect on increasing pedestrian counts can be seen on
Saturday in February at the State Library.}\label{fig:facet-time}
\end{figure}
\end{CodeChunk}

The work is inspired by \citet{Wickham2012glyph}, which uses linear
algebra to display spatio-temporal data as glyphs on maps. It is also
related to recent work by \citet{R-geofacet} which provides methods in
the \pkg{geofacet} package to arrange data plots into a grid, while
preserving the geographical position. Both of these show data in a
spatial context.

In contrast, calendar-based graphics unpack the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example,
\citet{VanWijkCluster1999} developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colors indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasonal variation such as school vacations, all of which have
influence over the turn-outs in the office. The calendar-based heatmap
was implemented in two R packages: \pkg{ggTimeSeries}
\citep{R-ggTimeSeries} and \pkg{ggcal} \citep{R-ggcal}. However, these
techniques are limited to color-encoding graphics and are unable to use
time scales smaller than a day. Time of day, which serves as one of the
most important aspects in explaining substantial variations arising from
the pedestrian sensor data, will be neglected through daily aggregation.
Additionally, if simply using colored blocks rather than curves, it may
become perceptually difficult to estimate the shape positions and
changes, although using curves comes with the cost of more display
capacity \citep{cleveland1984graphical, lam2007overview}.

We propose a new algorithm to go beyond the calendar-based heatmap. The
approach is developed with three conditions in mind: (1) to display
time-of-day variation in addition to longer temporal components such as
day-of-week and day-of-year; (2) to incorporate line graphs and other
types of glyphs into the graphical toolkit for the calendar layout; (3)
to enable overlaying plots consisting of multiple time series. The
proposed algorithm has been implemented in the \code{frame_calendar}
function in the \pkg{sugrrants} package \citep{R-sugrrants} using
\proglang{R} \citep{R-base}.

The remainder of the paper is organized as follows. Section
\ref{sec:algorithm} demonstrates the construction of the calendar layout
in depth. Section \ref{sec:opt} lists and describes the options that
come with the \code{frame_calendar} function. Section \ref{sec:examples}
presents some variations of its usage. Section \ref{sec:discussion}
discusses the advantages and disadvantages of the method.

\section{Construction}\label{construction}

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year 2016. This is achieved by the
\code{frame_calendar} function, which computes the coordinates on the
calendar for the input data variables. These can then be plotted using
the usual \pkg{ggplot2} package \citep{R-ggplot2} functions. All of the
grammar of graphics \citep{wilkinson2006grammar, wickham2009ggplot2} can
be applied.

\begin{CodeChunk}
\begin{figure}

@@

\caption[The calendar-based display of hourly foot
traffic at Flinders Street Station using line glyphs. The arrangement of
the data into a \(3 \times 4\) monthly grid represents all the traffic
in 2016. The disparities between week day and weekend along with public
holiday are immediately apparent.]{The calendar-based display of hourly foot
traffic at Flinders Street Station using line glyphs. The arrangement of
the data into a \(3 \times 4\) monthly grid represents all the traffic
in 2016. The disparities between week day and weekend along with public
holiday are immediately apparent.}\label{fig:flinders-2016}
\end{figure}
\end{CodeChunk}

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
\citep{Wickham2012glyph}. To make a year long calendar requires cells
for days, embedded in blocks corresponding to months, organized into a
grid layout for a year. Each month can be captured with 35 (5 \(\times\)
7) cells, where the top left is Monday of week 1, and the bottom right
is Sunday of week 5 by default. These cells provide a micro canvas on
which to plot the data. The first day of the month could be any of
Monday--Sunday, which is determined by the year of the calendar. Months
are of different lengths, ranging from 28 to 31 days, and each month
could extend over six weeks but the convention in these months is to
wrap the last few days up to the top row of the block. The notation for
creating these cells is as follows:

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=360pt,height=250pt]{figure/month}

@@

\caption[Illustration of the indexing layout for cells in
a month, where \(k\) is day of the week, \(g\) is day of the month,
\((i, j)\) indicates grid position.]{Illustration of the indexing layout for cells in
a month, where \(k\) is day of the week, \(g\) is day of the month,
\((i, j)\) indicates grid position.}\label{fig:month-diagram}
\end{figure}
\end{CodeChunk}

@@

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=360pt,height=250pt]{figure/year-diagram}

@@

\caption[Illustration of the indexing layout for months of
one year, where \(M\) and \(N\) indicate number of rows and columns,
\(b\) is a space parameter separating cells.]{Illustration of the indexing layout for months of
one year, where \(M\) and \(N\) indicate number of rows and columns,
\(b\) is a space parameter separating cells.}\label{fig:year-diagram}
\end{figure}
\end{CodeChunk}

@@

\section{Options}\label{options}

@@

The algorithm has several optional parameters that modify the layout,
direction of display, scales, plot size and switching to polar
coordinates. These are accessible to the user by the inputs to the
function \code{frame_calendar}:

@@

It is assumed that the \texttt{data} is in tidy format
\citep{wickham2014tidy}, and \code{x}, \code{y} are the variables that
will be mapped to the horizontal and vertical axes in each cell. For
example, the \code{x} is the time of the day, and \code{y} is the count
(Figure \ref{fig:flinders-2016}). The \code{date} argument specifies the
date variable used to construct the calendar layout.

@@

\subsection{Layouts}\label{layouts}

@@

\subsection{Polar transformation}\label{polar-transformation}

When \code{polar = TRUE}, a polar transformation is carried out on the
data. The computation is similar to the one described in
\citet{Wickham2012glyph}. Figure \ref{fig:flinders-polar} shows star
plots embedded in the monthly calendar layout, which is equivalent to
Figure \ref{fig:flinders-2016} placed in polar coordinates.

\begin{CodeChunk}
\begin{figure}

@@

\caption[Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The daily periodicity on
work days are clearly visible.]{Figure \ref{fig:flinders-2016} in circular
layout, which is referred to as star plots. The daily periodicity on
work days are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{CodeChunk}

\subsection{Scales}\label{scales}

By default, global scaling is done for values in each plot, with the
global minimum and maximum used to fit values into each cell. If the
emphasis is comparing trend rather than magnitude, it is useful to scale
locally. For temporal data this would harness the temporal components.
The choices include: free scale within each cell (\code{free}), cells
derived from the same day of the week (\code{free_wday}), or cells from
the same day of the month (\code{free_mday}). The scaling allows for the
comparisons of absolute or relative values, and the emphasis of
different temporal variations.

@@

The \code{free_wday} scales each week day together. It can be useful to
comparing trends across week days, allowing relative patterns for
weekends versus week days to be examined. Similarly, the
\code{free_mday} uses free scaling for any day within a given month.

\begin{CodeChunk}
\begin{figure}

@@

\caption[Line glyphs on the calendar format showing
hourly foot traffic at Flinders Street Station, scaled over all the
days. The individual shape on a single day becomes more distinctive,
however it is impossible to compare the size of peaks between days.]{Line glyphs on the calendar format showing
hourly foot traffic at Flinders Street Station, scaled over all the
days. The individual shape on a single day becomes more distinctive,
however it is impossible to compare the size of peaks between days.}\label{fig:flinders-free}
\end{figure}
\end{CodeChunk}

\subsection{Orientation}\label{orientation}

@@

\subsection{Language support}\label{language-support}

Most countries have adopted this western calendar layout, while the
languages used for week day and month would be different across
countries. We also offer languages other than English for text
labelling. Figure \ref{fig:chn} shows the same plot as Figure
\ref{fig:boxplot} labelled using simplified Chinese characters.

\section{Variations}\label{variations}

@@

\subsection{Overlaying and faceting
subsets}\label{overlaying-and-faceting-subsets}

@@

\caption[Overlaying line graphs of the 3 sensors in the monthly
calendar. Flagstaff station is not as busy as the other two on non-work
days.]{Overlaying line graphs of the 3 sensors in the monthly
calendar. Flagstaff station is not as busy as the other two on non-work
days.}\label{fig:overlay}
\end{figure}
\end{CodeChunk}

@@

\caption[Line charts, embedded in the \(6 \times 2\) monthly
calendar, colored and faceted by the 3 sensors. The variations of an
individual sensor are emphasised, and the shapes can be compared across
the cells and sensors.]{Line charts, embedded in the \(6 \times 2\) monthly
calendar, colored and faceted by the 3 sensors. The variations of an
individual sensor are emphasised, and the shapes can be compared across
the cells and sensors.}\label{fig:facet}
\end{figure}
\end{CodeChunk}

\subsection{Different types of plots}\label{different-types-of-plots}

The \code{frame_calendar} function is not constrained to line plots. The
full range of plotting capabilities in \pkg{ggplot2} is essentially
available. Figure \ref{fig:scatterplot} shows a lag scatterplot at
Flinders Street Station, where the current hourly count is assigned to
the \code{x} argument and the lagged hourly count to the \code{y}
argument. This figure is organized in the daily calendar layout. Figure
\ref{fig:scatterplot} indicates two primary patterns, strong
autocorrelation on weekends, and weaker autocorrelation on work days. At
the higher counts, on week days, the next hour sees possibly substantial
increase or decrease in counts, essentially revealing a bimodal
distribution of consecutive counts, as supported by Figure
\ref{fig:flinders-2016}.

\begin{CodeChunk}
\begin{figure}

@@

\caption[Lag scatterplot in the daily calendar layout.
Previous hour's count is plotted against each hour's count at Flinders
Street Station to demonstrate the autocorrelation at lag 1. The
correlation between them is more consistent on non-work days than work
days.]{Lag scatterplot in the daily calendar layout.
Previous hour's count is plotted against each hour's count at Flinders
Street Station to demonstrate the autocorrelation at lag 1. The
correlation between them is more consistent on non-work days than work
days.}\label{fig:scatterplot}
\end{figure}
\end{CodeChunk}

@@

\caption[Side-by-side boxplots of hourly counts for all the 43
sensors in December 2016, with the loess smooth line superimposed on
each day. It shows the hourly distribution in the city as a whole. There
is one sensor attracting a larger number of people on New Year's Eve
than the rest.]{Side-by-side boxplots of hourly counts for all the 43
sensors in December 2016, with the loess smooth line superimposed on
each day. It shows the hourly distribution in the city as a whole. There
is one sensor attracting a larger number of people on New Year's Eve
than the rest.}\label{fig:boxplot}
\end{figure}
\end{CodeChunk}

\begin{CodeChunk}
\begin{figure}

{\centering \includegraphics[width=\textwidth]{figure/chn-1}

@@

\caption[The same plot as Figure \ref{fig:boxplot}, but with the
month and week day labels in Chinese. It demonstrates the natural
support for languages other than English.]{The same plot as Figure \ref{fig:boxplot}, but with the
month and week day labels in Chinese. It demonstrates the natural
support for languages other than English.}\label{fig:chn}
\end{figure}
\end{CodeChunk}

\section{Discussion}\label{discussion}

@@

\bibliography{packages.bib,references.bib}

\end{document}

>>>>

% !TeX root = RJwrapper.tex
\title{Calendar-based graphics for visualizing people's daily schedules}
\author{by Earo Wang, Dianne Cook, Rob J Hyndman}

\maketitle

\abstract{%
Calendars are broadly used in society to display temporal information,
and events. This paper describes a new R package with functionality to
organize and display temporal data, collected on sub-daily resolution,
into a calendar layout. The function \code{frame\_calendar} uses linear
algebra on the date variable to restructure data into a format lending
itself to calendar layouts. The user can apply the grammar of graphics
to create plots inside each calendar cell, and thus the displays
synchronize neatly with \pkg{ggplot2} graphics. The motivating
application is studying pedestrian behavior in Melbourne, Australia,
based on counts which are captured at hourly intervals by sensors
scattered around the city. Faceting by the usual features such as day
and month, was insufficient to examine the behavior. Making displays on
a monthly calendar format helps to understand pedestrian patterns
relative to events such as work days, weekends, holidays, and special
events. The layout algorithm has several format options and variations.
It is implemented in the R package \pkg{sugrrants}.
}

\hypertarget{introduction}{%
\section{Introduction}\label{introduction}}

We develop a method for organizing and visualizing temporal data,
collected at sub-daily intervals, into a calendar layout. The calendar
format is created using linear algebra, giving a restructuring of the
data, that can then be piped into grammar of graphics definitions of
plots, as used in \pkg{ggplot2} (Wickham et al. 2018). The data
restructuring approach is consistent with the tidy data principles
available in the \pkg{tidyverse} (Wickham 2017) suite. The methods are
implemented in a new package called \pkg{sugrrants} (Wang, Cook, and
Hyndman 2018) .

The purpose of the calendar-based visualization is to provide insights
into people's daily schedules relative to events such as work days,
weekends, holidays, and special events. This work was originally
motivated by studying foot traffic in the city of Melbourne, Australia
(City of Melbourne 2017). There have been 43 sensors installed that
count pedestrians every hour across the downtown area (Figure
\ref{fig:ped-map}). The dataset can shed light on people's daily
rhythms, and assist the city administration and local businesses with
event planning and operational management. A routine examination of the
data would involve constructing conventional time series plots to catch
a glimpse of temporal patterns. The faceted plots in Figure
\ref{fig:time-series-plot}, give an overall picture of the foot traffic
at 3 different sensors over 2016. Further faceting by day of the week
(Figure \ref{fig:facet-time}) provides a better glimpse of the daily and
sub-daily pedestrian patterns. The sensor data, like many temporal
datasets on human behaviors, lends itself to a number of exploratory
data visualization challenges:

@@

\begin{Schunk}
\begin{figure}

@@

\caption[Map of the Melbourne city area with purple dots indicating sensor locations]{Map of the Melbourne city area with purple dots indicating sensor locations.}\label{fig:ped-map}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}

@@

\caption[Time series plots showing the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne]{Time series plots showing the number of pedestrians in 2016 measured at 3 different sensors in the city of Melbourne. Colored by the sensors, small multiples of lines show that the foot traffic varies from one sensor to another in terms of both time and number. The weekly patterns look distinctive across these 3 sensors. There is an eye-catching spike which occurred at the State Library, caused by the annual White Night event on 20th of February.}\label{fig:time-series-plot}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}

@@

\caption[Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines]{Hourly pedestrian counts for 2016 faceted by sensors and days of the week using lines. It features at least two types of seasons---time of day and day of week---across all the sensors. The White Night effect on increasing pedestrian counts can be seen on Saturday in February at the State Library.}\label{fig:facet-time}
\end{figure}
\end{Schunk}

The work is inspired by Wickham et al. (2012), which uses linear algebra
to display spatio-temporal data as glyphs on maps. It is also related to
recent work by Hafen (2018) which provides methods in the \pkg{geofacet}
package to arrange data plots into a grid, while preserving the
geographical position. Both of these show data in a spatial context.

In contrast, calendar-based graphics unpack the temporal variable, at
different resolutions, to digest multiple seasonalities, and special
events. There is some existing work in this area. For example, Van Wijk
and Van Selow (1999) developed a calendar view of the heatmap to
represent the number of employees in the work place over a year, where
colors indicate different clusters derived from the days. It contrasts
week days and weekends, highlights public holidays, and presents other
known seasonal variation such as school vacations, all of which have
influence over the turn-outs in the office. The calendar-based heatmap
was implemented in two R packages: \pkg{ggTimeSeries} (Kothari and Ather
2016) and \pkg{ggcal} (Jacobs 2017). However, these techniques are
limited to color-encoding graphics and are unable to use time scales
smaller than a day. Time of day, which serves as one of the most
important aspects in explaining substantial variations arising from the
pedestrian sensor data, will be neglected through daily aggregation.
Additionally, if simply using colored blocks rather than curves, it may
become perceptually difficult to estimate the shape positions and
changes, although using curves comes with the cost of more display
capacity (Cleveland and McGill 1984; Lam, Munzner, and Kincaid 2007).

We propose a new algorithm to go beyond the calendar-based heatmap. The
approach is developed with three conditions in mind: (1) to display
time-of-day variation in addition to longer temporal components such as
day-of-week and day-of-year; (2) to incorporate line graphs and other
types of glyphs into the graphical toolkit for the calendar layout; (3)
to enable overlaying plots consisting of multiple time series. The
proposed algorithm has been implemented in the \code{frame\_calendar}
function in the \pkg{sugrrants} package using R.

The remainder of the paper is organized as follows. Section
\ref{sec:algorithm} demonstrates the construction of the calendar layout
in depth. Section \ref{sec:opt} lists and describes the options that
come with the \code{frame\_calendar} function. Section
\ref{sec:examples} presents some variations of its usage. Section
\ref{sec:discussion} discusses the advantages and disadvantages of the
method.

\hypertarget{construction}{%
\section{Construction}\label{construction}}

@@

Figure \ref{fig:flinders-2016} shows the line glyphs framed in the
monthly calendar over the year 2016. This is achieved by the
\code{frame\_calendar} function, which computes the coordinates on the
calendar for the input data variables. These can then be plotted using
the usual \pkg{ggplot2} package (Wickham et al. 2018) functions. All of
the grammar of graphics (Wilkinson 2005; Wickham 2009) can be applied.

\begin{Schunk}
\begin{figure}

@@

\caption[The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs]{The calendar-based display of hourly foot traffic at Flinders Street Station using line glyphs. The arrangement of the data into a $3  imes 4$ monthly grid represents all the traffic in 2016. The disparities between week day and weekend along with public holiday are immediately apparent.}\label{fig:flinders-2016}
\end{figure}
\end{Schunk}

The algorithm for constructing a calendar plot uses linear algebra,
similar to that used in the glyph map displays for spatio-temporal data
(Wickham et al. 2012). To make a year long calendar requires cells for
days, embedded in blocks corresponding to months, organized into a grid
layout for a year. Each month can be captured with 35 (5 \(\times\) 7)
cells, where the top left is Monday of week 1, and the bottom right is
Sunday of week 5 by default. These cells provide a micro canvas on which
to plot the data. The first day of the month could be any of
Monday--Sunday, which is determined by the year of the calendar. Months
are of different lengths, ranging from 28 to 31 days, and each month
could extend over six weeks but the convention in these months is to
wrap the last few days up to the top row of the block. The notation for
creating these cells is as follows:

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=360pt,height=250pt]{img/month}

@@

\caption[Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $(i, j)$ indicates grid position]{Illustration of the indexing layout for cells in a month, where $k$ is day of the week, $g$ is day of the month, $(i, j)$ indicates grid position.}\label{fig:month-diagram}
\end{figure}
\end{Schunk}

@@

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=360pt,height=250pt]{img/year-diagram}

@@

\caption[Illustration of the indexing layout for months of one year, where $M$ and $N$ indicate number of rows and columns, $b$ is a space parameter separating cells]{Illustration of the indexing layout for months of one year, where $M$ and $N$ indicate number of rows and columns, $b$ is a space parameter separating cells.}\label{fig:year-diagram}
\end{figure}
\end{Schunk}

@@

\hypertarget{options}{%
\section{Options}\label{options}}

@@

The algorithm has several optional parameters that modify the layout,
direction of display, scales, plot size and switching to polar
coordinates. These are accessible to the user by the inputs to the
function \code{frame\_calendar}:

\newpage

@@

It is assumed that the \texttt{data} is in tidy format (Wickham 2014),
and \code{x}, \code{y} are the variables that will be mapped to the
horizontal and vertical axes in each cell. For example, the \code{x} is
the time of the day, and \code{y} is the count (Figure
\ref{fig:flinders-2016}). The \code{date} argument specifies the date
variable used to construct the calendar layout.

@@

\hypertarget{layouts}{%
\subsection{Layouts}\label{layouts}}

@@

\hypertarget{polar-transformation}{%
\subsection{Polar transformation}\label{polar-transformation}}

When \code{polar = TRUE}, a polar transformation is carried out on the
data. The computation is similar to the one described in Wickham et al.
(2012). Figure \ref{fig:flinders-polar} shows star plots embedded in the
monthly calendar layout, which is equivalent to Figure
\ref{fig:flinders-2016} placed in polar coordinates.

\begin{Schunk}
\begin{figure}

@@

\caption{Figure \ref{fig:flinders-2016} in circular layout, which is referred to as star plots. The daily periodicity on work days are clearly visible.}\label{fig:flinders-polar}
\end{figure}
\end{Schunk}

\hypertarget{scales}{%
\subsection{Scales}\label{scales}}

By default, global scaling is done for values in each plot, with the
global minimum and maximum used to fit values into each cell. If the
emphasis is comparing trend rather than magnitude, it is useful to scale
locally. For temporal data this would harness the temporal components.
The choices include: free scale within each cell (\code{free}), cells
derived from the same day of the week (\code{free\_wday}), or cells from
the same day of the month (\code{free\_mday}). The scaling allows for
the comparisons of absolute or relative values, and the emphasis of
different temporal variations.

@@

The \code{free\_wday} scales each week day together. It can be useful to
comparing trends across week days, allowing relative patterns for
weekends versus week days to be examined. Similarly, the
\code{free\_mday} uses free scaling for any day within a given month.

\begin{Schunk}
\begin{figure}

@@

\caption[Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days]{Line glyphs on the calendar format showing hourly foot traffic at Flinders Street Station, scaled over all the days. The individual shape on a single day becomes more distinctive, however it is impossible to compare the size of peaks between days.}\label{fig:flinders-free}
\end{figure}
\end{Schunk}

\hypertarget{orientation}{%
\subsection{Orientation}\label{orientation}}

@@

\hypertarget{language-support}{%
\subsection{Language support}\label{language-support}}

Most countries have adopted this western calendar layout, while the
languages used for week day and month would be different across
countries. We also offer languages other than English for text
labelling. Figure \ref{fig:chn-embedded} shows the same plot as Figure
\ref{fig:boxplot} labelled using simplified Chinese characters.

\hypertarget{variations}{%
\section{Variations}\label{variations}}

@@

\hypertarget{overlaying-and-faceting-subsets}{%
\subsection{Overlaying and faceting
subsets}\label{overlaying-and-faceting-subsets}}

@@

\caption[Overlaying line graphs of the 3 sensors in the monthly calendar]{Overlaying line graphs of the 3 sensors in the monthly calendar. Flagstaff station is not as busy as the other two on non-work days.}\label{fig:overlay}
\end{figure}
\end{Schunk}

@@

\caption[Line charts, embedded in the $6    imes 2$ monthly calendar, colored and faceted by the 3 sensors]{Line charts, embedded in the $6     imes 2$ monthly calendar, colored and faceted by the 3 sensors. The variations of an individual sensor are emphasised, and the shapes can be compared across the cells and sensors.}\label{fig:facet}
\end{figure}
\end{Schunk}

\hypertarget{different-types-of-plots}{%
\subsection{Different types of plots}\label{different-types-of-plots}}

The \code{frame\_calendar} function is not constrained to line plots.
The full range of plotting capabilities in \pkg{ggplot2} is essentially
available. Figure \ref{fig:scatterplot} shows a lag scatterplot at
Flinders Street Station, where the current hourly count is assigned to
the \code{x} argument and the lagged hourly count to the \code{y}
argument. This figure is organized in the daily calendar layout. Figure
\ref{fig:scatterplot} indicates two primary patterns, strong
autocorrelation on weekends, and weaker autocorrelation on work days. At
the higher counts, on week days, the next hour sees possibly substantial
increase or decrease in counts, essentially revealing a bimodal
distribution of consecutive counts, as supported by Figure
\ref{fig:flinders-2016}.

\begin{Schunk}
\begin{figure}

@@

\caption[Lag scatterplot in the daily calendar layout]{Lag scatterplot in the daily calendar layout. Previous hour's count is plotted against each hour's count at Flinders Street Station to demonstrate the autocorrelation at lag 1. The correlation between them is more consistent on non-work days than work days.}\label{fig:scatterplot}
\end{figure}
\end{Schunk}

@@

\caption[Side-by-side boxplots of hourly counts for all the 43 sensors in December 2016, with the loess smooth line superimposed on each day]{Side-by-side boxplots of hourly counts for all the 43 sensors in December 2016, with the loess smooth line superimposed on each day. It shows the hourly distribution in the city as a whole. There is one sensor attracting a larger number of people on New Year's Eve than the rest.}\label{fig:boxplot}
\end{figure}
\end{Schunk}

\begin{Schunk}
\begin{figure}

{\centering \includegraphics[width=\textwidth]{img/chn-1-embedded}

@@

\caption{The same plot as Figure \ref{fig:boxplot}, but with the month and week day labels in Chinese. It demonstrates the natural support for languages other than English.}\label{fig:chn-embedded}
\end{figure}
\end{Schunk}

\hypertarget{discussion}{%
\section{Discussion}\label{discussion}}

@@

\hypertarget{refs}{}
\leavevmode\hypertarget{ref-ped}{}%
City of Melbourne. 2017. \emph{Pedestrian Volume in Melbourne}. City of
Melbourne, Australia. \url{http://www.pedestrian.melbourne.vic.gov.au}.

\leavevmode\hypertarget{ref-cleveland1984graphical}{}%
Cleveland, William S, and Robert McGill. 1984. ``Graphical Perception:
Theory, Experimentation, and Application to the Development of Graphical
Methods.'' \emph{Journal of the American Statistical Association} 79
(387). Taylor \& Francis: 531--54.

\leavevmode\hypertarget{ref-R-geofacet}{}%
Hafen, Ryan. 2018. \emph{Geofacet: 'Ggplot2' Faceting Utilities for
Geographical Data}. \url{https://CRAN.R-project.org/package=geofacet}.

\leavevmode\hypertarget{ref-R-ggcal}{}%
Jacobs, Jay. 2017. \emph{Ggcal: Calendar Plot Using Ggplot2}.
\url{https://github.com/jayjacobs/ggcal}.

\leavevmode\hypertarget{ref-R-ggTimeSeries}{}%
Kothari, Aditya, and Ather. 2016. \emph{ggTimeSeries: Nicer Time Series
Visualisations with Ggplot Syntax.}
\url{https://github.com/Ather-Energy/ggTimeSeries}.

\leavevmode\hypertarget{ref-lam2007overview}{}%
Lam, Heidi, Tamara Munzner, and Robert Kincaid. 2007. ``Overview Use in
Multiple Visual Information Resolution Interfaces.'' \emph{IEEE
Transactions on Visualization and Computer Graphics} 13 (6). IEEE:
1278--85.

\leavevmode\hypertarget{ref-VanWijkCluster1999}{}%
Van Wijk, Jarke J., and Edward R. Van Selow. 1999. ``Cluster and
Calendar Based Visualization of Time Series Data.'' In \emph{Information
Visualization, 1999. INFOVIS 1999 Proceedings. IEEE Symposium on}, 4--9.
IEEE.

\leavevmode\hypertarget{ref-R-sugrrants}{}%
Wang, Earo, Di Cook, and Rob Hyndman. 2018. \emph{Sugrrants: Supporting
Graphs for Analysing Time Series}. \url{https://pkg.earo.me/sugrrants}.

\leavevmode\hypertarget{ref-wickham2009ggplot2}{}%
Wickham, Hadley. 2009. \emph{Ggplot2: Elegant Graphics for Data
Analysis}. New York, NY: Springer-Verlag New York.

\leavevmode\hypertarget{ref-wickham2014tidy}{}%
---------. 2014. ``Tidy Data.'' \emph{Journal of Statistical Software}
59 (10). Foundation for Open Access Statistics: 1--23.

\leavevmode\hypertarget{ref-R-tidyverse}{}%
---------. 2017. \emph{Tidyverse: Easily Install and Load the
'Tidyverse'}. \url{https://CRAN.R-project.org/package=tidyverse}.

\leavevmode\hypertarget{ref-R-ggplot2}{}%
Wickham, Hadley, Winston Chang, Lionel Henry, Thomas Lin Pedersen,
Kohske Takahashi, Claus Wilke, and Kara Woo. 2018. \emph{Ggplot2: Create
Elegant Data Visualisations Using the Grammar of Graphics}.

\leavevmode\hypertarget{ref-Wickham2012glyph}{}%
Wickham, Hadley, Heike Hofmann, Charlotte Wickham, and Dianne Cook.
2012. ``Glyph-Maps for Visually Exploring Temporal Patterns in Climate
Data and Models.'' \emph{Environmetrics} 23 (5): 382--93.

\leavevmode\hypertarget{ref-wilkinson2006grammar}{}%
Wilkinson, Leland. 2005. \emph{The Grammar of Graphics (Statistics and
Computing)}. Secaucus, NJ: Springer-Verlag New York, Inc.

\bibliography{references.bib,packages.bib}

\address{%
Earo Wang\\
Monash University\\
Department of Econometrics and Business Statistics,\\ Monash University, VIC 3800\\ Australia\\
}
\href{mailto:earo.wang@monash.edu}{\nolinkurl{earo.wang@monash.edu}}

\address{%
Dianne Cook\\
Monash University\\
Department of Econometrics and Business Statistics,\\ Monash University, VIC 3800\\ Australia\\
}
\href{mailto:dicook@monash.edu}{\nolinkurl{dicook@monash.edu}}

\address{%
Rob J Hyndman\\
Monash University\\
Department of Econometrics and Business Statistics,\\ Monash University, VIC 3800\\ Australia\\
}
\href{mailto:rob.hyndman@monash.edu}{\nolinkurl{rob.hyndman@monash.edu}}
